---
title: ""
author: ""
date: ""
output: pdf_document
header-includes:
  - \usepackage{amsmath}
  - \usepackage{bm}
geometry: margin=1in
fontsize: 12pt
linestretch: 1
---
**Final Project – EDA & Preliminary Results**  
Claire Yang, Saman De Silva, & Hans Elasri  
`r format(Sys.Date())`
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# libraries for the project are here
library(tidyverse)
library(lme4)
library(broom.mixed)
library(ggplot2)
library(knitr)
library(readr)
library(broom.mixed)
library(gt)
library(dplyr)
```
```{r}
# loading in cleaned data
dat <- read_csv("data/cfps2022_clean_persons.csv")
```
## 1. Background & Primary Research Questions
In this project, we use the 2022 wave of the China Family Panel Survey (CFPS) to study how adults' perceived income satisfaction, perceived income relative to others, and perceived social status are related to depressive symptoms, as measured by the CES-D8 scale. Our primary research questions are: (1) Are higher levels of subjective socioeconomic status associated with lower depression scores, after adjusting for demographic and household characteristics, and if so is there an interactive effect between the subjective scores? (2) How much of the variation in depression occurs between individuals versus between families, communities, and counties? and (3) To what extent to these associations remain after controlling for proxies of socioeconomic status such as education, employment status, and household composition?

## 2. Data Sources & Structure
Our data come from the China Family Panel Study (CFPS) 2022, a nationally representative household survey administered by the Institute of Social Science Survey at Peking University. The raw person-level dataset contains 27,001 individuals, each linked to hierarchical identifiers for families, communities, counties, and providences. After recoding special missing codes and cleaning key SES and demographic variables, removing structurally invalid IDs, our dataset reduces to 23,727 individuals nested in 9,663 families, 3,220 communities, 760 counties, and 31 provinces. For the analytic sample, we drop respondents missing any of the three SES predictors of interest, missing depression scores, or under age 18. This produces a final dataset of 14,575 adults nested in 8,123 families, 2,856 communities, 702 counties, and all 31 providences. We believe that missingness is largely missing completely at random based on the user guide, but more work is needed to verify this assumption. All variables come from survey interviers (no administrative data or observational data) and we created no synthetic measures other than centering continuous covariates and contructing binary indicators. The major data decisions made were (1) treating CFPS special codes as true missingness, (2) dropping households with invalid/missing geographic IDs, and (3) restricting to respondents with complete SES and outcome information. These steps ensure comparability across respondents and preserve the multilevel structure needed for our 4 level model.


## 3. EDA Plots


## 4. Variance Decomposition


## 5. Data Trends


## 6. Mathmatical Model
Our model has four clustered levels: 
$i$ = individuals, 
$h$ = households/families, 
$c$ = communities, 
$k$ = counties.
Our primary outcome variable is a (0-24) depression score: $Y_{ihck}$. \newline
There are three primary predictors of interest (all grand mean centered Likert scores):
\begin{itemize}
  \item Income Satisfaction: $IS_{ij}^* = income\_satis_{ij} - \overline{income\_satis}$
  \item Perception of Relative Income: $IR_{ij}^* = income\_rel_{ij} - \overline{income\_rel}$
  \item Perception of Relative Social Status: $IS_{ij}^* = status\_rel{ij} - \overline{status\_rel}$
\end{itemize}
\begin{itemize}
  \item Let $\bm{X}_{ihck}$ be a column vector of control variables, mostly level 1 (e.g. age, gender, ethnicity, rural/urban, CCP party member, retired, pension, etc)
  \item Let $\bm{\delta}$ is the corresponding vector of fixed effect coefficients
\end{itemize}
The multilevel model is then as follows:\newline
Level 1 (Individual):
$$
Y_{ihck} = \beta_{0hck} + \beta_1IS_{ihck}^* + \beta_2IR^*_{ihck}+\beta_3SR^*_{ihck} +\mathbf{X}^{T}_{ihck}\bm{\delta} + \sum_{p=2}^P \beta_p^{(prov)}D_{p(ihck)} + \varepsilon_{ihck}, \quad \varepsilon_{ihck} \sim \mathcal{N}(0,\sigma^2)
$$
where $D_{p(ihck)}$ is a dummy indicating that the individaul is in providence $p$.  \newline
Level 2 (Family):
$$
\beta_{0hck} = \gamma_{00ck}+u_{0hck},\quad u_{0hck}\sim \mathcal{N}(n,\tau^2_{family})
$$
Level 3 (Community):
$$
\beta_{00ck} = \gamma_{000k}+v_{0ck},\quad v_{0ck}\sim \mathcal{N}(n,\tau^2_{community})
$$
Level 4 (County):
$$
\beta_{000k} = \gamma_{0000}+w_{0k},\quad w_{0k}\sim \mathcal{N}(n,\tau^2_{county})
$$
Combined:
$$
Y_{ihck}
= \gamma_{0000}
+ \beta_1\,\text{IS}^*_{ihck}
+ \beta_2\,\text{IR}^*_{ihck}
+ \beta_3\,\text{SR}^*_{ihck}
+ \mathbf{X}_{ihck}^\top \boldsymbol{\delta}
+ \sum_{p=2}^{P} \beta_{p}^{(\text{prov})} D_{p(ihck)}
+ w_{0k} + v_{0ck} + u_{0hck}
+ \varepsilon_{ihck}.
$$

## 7. Initial Findings

```{r}
# -------------------------------------------
# Center SES predictors and continuous controls
# -------------------------------------------

clean_dat <- dat %>%
  mutate(
    # SES predictors (grand-mean centered)
    c_income_satis = scale(income_satis, center = TRUE, scale = FALSE),
    c_income_rel   = scale(income_rel, center = TRUE, scale = FALSE),
    c_status_rel   = scale(status_rel, center = TRUE, scale = FALSE),

    # controls (center continuous)
    c_age          = scale(age, center = TRUE, scale = FALSE),
    c_num_child    = scale(num_children_u16, center = TRUE, scale = FALSE),
    c_family_size  = scale(family_size, center = TRUE, scale = FALSE),
    c_wordlist     = scale(wordlist_raw, center = TRUE, scale = FALSE),
    c_mathlist     = scale(mathlist_raw, center = TRUE, scale = FALSE),
  )

# -------------------------------------------
# Convert categorical controls to factors
# -------------------------------------------

clean_dat <- clean_dat %>%
  mutate(
    gender = factor(gender),
    employ_status = factor(employ_status),
    handedness = factor(handedness),
    province_id = factor(province_id)   # fixed effects
  )

# -------------------------------------------
# Fit the 4-level random intercept model
# -------------------------------------------

model_basic <- lmer(
  cesd8 ~ 
    # SES predictors
    c_income_satis + c_income_rel + c_status_rel +
    
    # Province fixed effects
    province_id +
    
    # Level-1 controls
    c_age + gender + ethnicity_han + hukou_rural + urban +
    party_member + retired + pension +
    employ_status + 
    c_num_child + c_family_size +
    c_wordlist + c_mathlist +
    smoke_age + handedness +
    
    # Random intercepts (4 levels)
    (1 | county_id) + (1 | community_id) + (1 | family_id),
  
  data = clean_dat,
  REML = TRUE
)
```
```{r, message=TRUE}
# Extract selected fixed effects and add p-values from t-stats
coef_table <- tidy(model_basic, effects = "fixed") %>%
  filter(term %in% c(
    "c_income_satis",
    "c_income_rel",
    "c_status_rel",
    "c_age",
    "gendermale"
  )) %>%
  mutate(
    term = recode(term,
      "c_income_satis" = "Income satisfaction (centered)",
      "c_income_rel"   = "Relative income (centered)",
      "c_status_rel"   = "Perceived social status (centered)",
      "c_age"          = "Age (centered)",
      "gendermale"     = "Male"
    ),
    # compute two-sided p-value from t statistic
    p_value = 2 * pnorm(abs(statistic), lower.tail = FALSE)
  ) %>%
  transmute(
    Term = term,
    Estimate = estimate,
    SE = std.error,
    `t stat` = statistic,
    `p-value` = p_value
  )

# Random-effect variances (no p-values)
re_table <- tidy(model_basic, effects = "ran_pars") %>%
  select(group, term, estimate) %>%
  filter(term == "sd__(Intercept)") %>%
  mutate(
    group = recode(group,
      "family_id"    = "Family SD",
      "community_id" = "Community SD",
      "county_id"    = "County SD",
      "Residual"     = "Residual SD"
    ),
    Term = group,
    Estimate = estimate,
    SE = NA_real_,
    `t stat` = NA_real_,
    `p-value` = NA_real_
  ) %>%
  select(Term, Estimate, SE, `t stat`, `p-value`)

# Combine into one table
final_table <- bind_rows(coef_table, re_table)

library(knitr)

# Make a simplified version of the table (round numbers, rename columns)
simple_table <- final_table %>%
  select(-`p-value`) %>% 
  mutate(
    Estimate = round(Estimate, 3),
    SE       = round(SE, 3),
    `t stat` = round(`t stat`, 3)
  )

kable(
  simple_table,
  booktabs = TRUE,
  caption = "Effects of Subjective SES on Depression (Multilevel Model)"
)

```

The cross-classified model estimates how perceived socioeconomic status measures relate to depression as measured by the CES-D8 survey among Chinese adults while controlling for demographic stats, household characteristics, and providence fixed effects. After adjusting for the covariates, all three predictors of relative socioeconomic measures are statistically significant at the 1\% level ($p < 0.001$). Income satisfaction shows the strongest association with depression: a one unit increase (relative to the sample mean) predicts a 0.53 point decrease in CES-D8 scores (this corresponds loosely to "half of a symptom"). Perceived relative income ($\beta = -0.17$) and perceived social status ($\beta=-0.33$) are also economically significant. Taken together, the findings suggest that individuals who feel better off or higher status than their peers tend to experience fewer depressive symptoms. Random-effects estimates show that most of the clustering occurs at the family level (var $\approx 2.31$) with smaller but still meaningful clustering happening at the community (var $\approx 0.29$) and county (var $\approx 0.20$) levels which suggests heterogeneity in depressive symptoms across families and communities. The results seem to indicate that subjective socioeconomic well being is a robust, consistent predictor of mental health in the CFPS 2022 data, but more work is needed to determine if these effects have any interactive effects between each other.

## 8. Next Steps
Our next steps are to refine the model and assess robustness. We plan to test interactions among the three subjective SES measures, examine subgroup differences (e.g., by gender or age), and compare our current four-level random-intercept model with alternatives such as cross-classified structures or limited random slopes. A remaining concern will be if adding complexity to the model will make estimates unstable or slow to fit. We would appreciate feedback on which extensions are most appropriate for this type of dataset and leverage the 4 levels in a more comprehensive way.