---
title: "PLEASE WRITE A TITLETitle of Your Project"
author: "Saman De Silva, Hans Elasri, Claire Yang"
output: pdf_document
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{setspace}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{amsmath, amssymb}
  - \usepackage{graphicx}
  - \usepackage{bm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.pos = "H"
)
```


# Introduction

Briefly motivate the topic and clearly state your **research question**.

* What is the substantive question?
* Why does it matter?
* What gap or puzzle are you addressing?

Keep this **tight** — not the bulk of the manuscript.

---

# Data

## 2.1 Data Description

Describe:

* Data source(s)
* Sample size
* Key units (e.g., individuals, schools, time points)
* Outcome variable
* Main predictors
* Structure (cross-sectional, longitudinal, hierarchical)

```{r}
# Example: glimpse(data)
# glimpse(my_data)
```

## 2.2 Descriptive Statistics (Table 1)

Include a **Table 1** summarizing the data.

```{r}
# Example table code
# library(dplyr)
# library(knitr)
# library(kableExtra)

# my_data %>%
#   summarize(...) %>%
#   kable(booktabs = TRUE, digits = 2) %>%
#   kable_styling(latex_options = "hold_position")
```



# Methods

Explain **how** you answer the research question.

Include:

* Modeling strategy
* Why multilevel modeling is appropriate (if applicable)
* Fixed vs random effects
* Key assumptions
* Estimation approach

Avoid dumping equations unless helpful — focus on interpretation.



# Model

asdf

---

# *Results

## 5.1 Main Results



## 5.2 Figures

**Figure 1**

```{r fig1, fig.cap="Figure 1: Descriptive or Model-Based Visualization"}
# plot(...)
```


## 5.3 Results Table



# Conclusion



# **Appendix A: Model Results (Required)**


## Model Call

```{r}
# lmer(
#   outcome ~ predictors + (1 | group),
#   data = my_data
# )
```

## Model Output

```{r}
# library(arm)
# display(model)
```



# **References**

```{r, results='asis', echo=FALSE}
# Example:
# cat("\\begin{thebibliography}{99}")
# cat("\\bibitem{ref1} Author. Title. Journal.")
# cat("\\end{thebibliography}")
```

