---
title: "PLEASE WRITE A TITLETitle of Your Project"
author: "Saman De Silva, Hans Elasri, Claire Yang"
output: pdf_document
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{setspace}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{amsmath, amssymb}
  - \usepackage{graphicx}
  - \usepackage{bm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.pos = "H"
)
```


# Introduction

China’s Zero-COVID strategy relied on aggressive non-pharmaceutical interventions, including lockdowns, mass testing, mobility restrictions, and community-level quarantines that varied in intensity across provinces and over time. The policies imposed substantial and uneven constraints on daily activity, social interaction, and employment providing a natural setting to study how large-scale and persistent containment measures may have affected mental health.

Assessing the psychological consequences of Zero-COVID is complicated by the simultaneous presence of COVID-19 morbidity and mortality, which may independently influence mental health. In addition, individuals differ in baseline levels of depression and vulnerability to stress, making cross-sectional comparisons difficult to interpret. Longitudinal data are therefore critical for isolating changes in mental health associated with evolving local conditions rather than stable individual differences.

In this project, we combine longitudinal data from the China Family Panel Studies (CFPS) with province-level measures of COVID-19 policy stringency from the Oxford COVID-19 Government Response Tracker and COVID-19 mortality data. Depression is measured using the CESD-20 scale. Our primary research question is whether increases in policy stringency and COVID-19 deaths are associated with changes in depressive symptoms.

We study this question using multilevel longitudinal models that exploit within-individual and across-province variation over time while accounting for the hierarchical structure of the CFPS sample. This approach allows us to estimate how changes in province policy stringency and pandemic intensity relate to changes in depression, net of persistent individual characteristics and common time trends.
---

# Data

\textbf{China Family Panel Studies (CFPS)}

Individual-level mental health and demographic data come from the China Family Panel Studies (CFPS), a nationally representative, biennial longitudinal survey administered by the Institute of Social Science Survey at Peking University. The CFPS tracks individuals, families, and communities across China and collects detailed information on health, economic conditions, education, and family structure.

We use four survey waves: 2016, 2018, 2020, and 2022. Depression is measured using the CESD-20 scale, which captures the frequency of depressive symptoms over the past week. The longitudinal structure of the CFPS allows us to observe changes in depressive symptoms within the same individuals before and during the COVID-19 period.

\textbf{COVID-19 Mortality Data}

Data on COVID-19 cases and deaths come from the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, sourced from official Chinese reporting channels including the National Health Commission of China and the China CDC. The dataset provides daily province-level counts of confirmed cases and deaths. We aggregate these data to the monthly level to match the temporal resolution of the survey data.

\textbf{Policy Stringency Data}

Measures of COVID-19 containment policy severity are drawn from the Oxford COVID-19 Government Response Tracker (OxCGRT). The OxCGRT provides systematic, cross-province indicators of government responses, including stay-at-home requirements, restrictions on gatherings, workplace closures, and travel controls. We use the Oxford Stringency Index, which summarizes the intensity of these policies on a standardized scale (Hale et al., 2021). Monthly averages of the stringency index are constructed to align with the CFPS survey timing.

\textbf{Data Construction}

All datasets are merged at the province–month level and aligned to individual CFPS interview months COVID-19 cases, deaths, and stringency measures are aggregated to monthly frequencies since we do not have the exact date of the interview. The final analytical dataset spans 2016–2022 and links individual-level mental health outcomes to time-varying local pandemic conditions and policy responses.


# Methods

Explain **how** you answer the research question.

Include:

* Modeling strategy
* Why multilevel modeling is appropriate (if applicable)
* Fixed vs random effects
* Key assumptions
* Estimation approach

Avoid dumping equations unless helpful — focus on interpretation.



# Model

asdf

---

# *Results

## 5.1 Main Results



## 5.2 Figures

**Figure 1**

```{r fig1, fig.cap="Figure 1: Descriptive or Model-Based Visualization"}
# plot(...)
```


## 5.3 Results Table



# Conclusion



# **Appendix A: Model Results (Required)**


## Model Call

```{r}
# lmer(
#   outcome ~ predictors + (1 | group),
#   data = my_data
# )
```

## Model Output

```{r}
# library(arm)
# display(model)
```



# **References**

```{r, results='asis', echo=FALSE}
# Example:
# cat("\\begin{thebibliography}{99}")
# cat("\\bibitem{ref1} Author. Title. Journal.")
# cat("\\end{thebibliography}")
```

